% Copyright 2021 Google LLC
%
% Use of this source code is governed by an MIT-style
% license that can be found in the LICENSE file or at
% https://opensource.org/licenses/MIT.

%!BIB program = biber
%!TeX program = lualatex
%!TeX spellcheck = en-US

\documentclass[hctr2.tex]{subfiles}
\begin{document}
\section{HCTR issues}
Our presentation of HCTR2 uses
names that differ from those used to present HCTR.

\begin{itemize}
    \item HCTR and HCTR2 use an unusual mode of stream encryption.
    \cite{hctr,hctrquad} refer to this mode as ``CTR mode'',
    but note the differences between this mode and 
    standard CTR mode\cite{ctr}.
    For the avoidance of ambiguity
    we name this mode \emph{XCTR mode}.
    \item What \cite{hctr,hctrquad} refer to as \(C, \mathit{CC}, D\)
    we refer to as \(U, \UU, V\)
    so that we can use \(P, C\) to refer to plaintext and ciphertext.
    We will use our names in what follows.
    \item Because of our use of POLYVAL, HCTR2 draws a distinction
    between the raw hash key \(\hgen\)
    and the value at which the polynomial is evaluated \(\hpoly\).
    HCTR has no such distinction
    and we use \(h\) in our discussion of HCTR.
\end{itemize}

Two errors in previous work on HCTR are addressed in HCTR2.

\subsection{Hash function}\label{badhash}
HCTR uses a hash function based on the polynomial
\begin{displaymath}
    H(X) = \poly(\pad(X) \Concat \fromint(\abs{X}) \Concat 0^n)
\end{displaymath}
Because it assumes a fixed-length tweak it simply sets \(X = M \Concat T\).
However, HCTR requires that the resulting polynomial be nonzero
even when \(X = \lambda\),
so as a special case \(H(\lambda) = h\).

Unfortunately, as \cite{kumarhctr} observes this is no longer
an injective map from \(X\)---we also have \(H(0) = h\). This
breaks the almost-XOR-universal property relied on
in the security bound and
straightforwardly leads to an attack in which two encryption queries
are presented, one of a block width, and the second extending
the first with a single zero (and assuming a zero length tweak).

\subsection{Security bound}\label{badproof}
HCTR was initially presented in \cite{hctr} with a security bound cubic
in the total size of all queries combined.  This is a little low
for comfort; if a 128-bit block cipher is used, it suggests a
distinguisher can be effective given tens of terabytes of queries,
which can quickly be reached over a fast link. 
\cite{hctrquad} presents a much more satisfactory quadratic bound,
but the proof has an error that invalidates the claimed bound.

While presented in a different way, 
the proof of~\cite{hctrquad} is very similar
to that of this paper: in the game RAND2, all queries get random
responses, the block cipher inputs are inferred, and a collision
in either the inferred plaintext or inferred ciphertext of the block cipher
sets the ``bad'' flag.  
Where HCTR2 has \(S = \MM \xor \UU \xor L\),
HCTR simply has \(S = \MM \xor \UU\). 
HCTR uses a fixed-length tweak;
for simplicity we assume a zero-length tweak in what follows.

For equation 17, the paper observes that the collision
\(Y_i^r \qeq \UU^s\) occurs iff \(h\) is one of the zeroes
of the polynomial \(Y_i^r \xor U^s \xor H(T^s, V^s)\).
This polynomial has degree at most
\(m^s\), and so can have
at most that many solutions. From this the paper infers
a quadratic bound on the probability of any such collision
given \(\sigma\) input blocks.

In equation 21, the paper considers collisions of the form 
\(S_i^r \qeq \MM^s\) and asserts that they are
quadratically bounded for the same reason. However this equation
is crucially different: unlike with \(Y_i^r\), the value we
infer for \(S_i^r\) depends on \(h\). The values of \(h\)
for which this collision occurs are
the zeroes of the polynomial
\begin{displaymath}
    M^r \xor H(T^r, N^r)
    \xor U^r \xor H(T^r, V^r) \xor \fromint(i) 
    \xor M^s \xor H(T^s, N^s)        
\end{displaymath}
This polynomial can have degree up to 
\(\max(m^r, m^s)\), and so the bound of
equation 17 does not apply.
If queries are permitted to be of any length,
this leads to a cubic security bound.
Consider an adversary who
sends a single query with \(x+1\) blocks, followed by
\(x\) queries of one block. For each \(i \in \{1\ldots x\}\) and
for each \(s \in \{2 \ldots x + 1\}\), we have at best
\(\prob{S_i^r = \MM^s} \leq (x + 1)/2^n\). Summing
all these bounds for each such pair, we find that
for queries with \(\sigma = 2x + 1\)
this technique yields an upper bound on 
the total probability of such a collision of
\((x^3 + x^2)/2^n\).

\cite{nandimail} observes that a quadratic bound can be recovered
if a bound \(l_\mathrm{max}\) is set on the maximum size of
a single query; in this case we can prove a bound
which is some small multiple of \(l_\mathrm{max}q\sigma/2^n\).
\end{document}
